---
title: "Mo_try_file"
author: "Tom"
date: "2026-04-09"
output: html_document
---

```{r, include=FALSE, echo=FALSE}
library(tidyverse)
library(patchwork)
library(ggplot2)
library(grid)
library(cowplot)
library(purrr)
```

# Load data
from now on we call this project mw (Moehne, Wester). We avoid mo as to not get into trouble with future Moehne projects
```{r, echo = FALSE}
# multiple runs from NED-flow by Kevin. So from oldest to most recent. We include these three because april contains microbes, june is used in thesis and september is the most recent one. 

mw_2504 <- read_tsv('data/beaver_250424_raw_taxonomy_data.tsv')
mw_2506 <- read_tsv('data/250627_taxon_out.tsv')
mw_2509 <- read_tsv('data/250910_taxa_out.tsv')

wcvp_names <- read.table("wcvp/wcvp_names.csv", sep="|", header=TRUE, quote = "", fill=TRUE, encoding = "UTF-8")
wcvp_dist <- read.table("wcvp/wcvp_distribution.csv", sep="|", header=TRUE, quote = "", fill=TRUE, encoding = "UTF-8")

```

# Add additional data
this data does not have db yet, so we assign kingdom group to see who's who.
Also we add layer names and ages per layer. 

```{r, echo = FALSE}
#--------------------------------- kingdom groups ------------------------------------------------
mw_phylum_groups <- tribble(
  ~phylum_name,           ~kingdom_group,
  # plants
  "Streptophyta",         "plants",
  "Chlorophyta",          "plants",
  "Rhodophyta",           "plants",
  
  # animals
  "Chordata",             "animals",
  "Arthropoda",           "animals",
  "Annelida",             "animals",
  "Mollusca",             "animals",
  "Cnidaria",             "animals",
  "Porifera",             "animals",
  "Nematoda",             "animals",
  "Rotifera",             "animals",
  
  # microbes (bacteria)
  "Acidobacteriota",      "microbes",
  "Actinomycetota",       "microbes",
  "Aquificota",           "microbes",
  "Armatimonadota",       "microbes",
  "Bacillota",            "microbes",
  "Bacteroidota",         "microbes",
  "Balneolota",           "microbes",
  "Bdellovibrionota",     "microbes",
  "Campylobacterota",     "microbes",
  "Chlamydiota",          "microbes",
  "Chlorobiota",          "microbes",
  "Chloroflexota",        "microbes",
  "Cyanobacteriota",      "microbes",
  "Deferribacterota",     "microbes",
  "Deinococcota",         "microbes",
  "Fibrobacterota",       "microbes",
  "Fusobacteriota",       "microbes",
  "Gemmatimonadota",      "microbes",
  "Ignavibacteriota",     "microbes",
  "Kiritimatiellota",     "microbes",
  "Lentisphaerota",       "microbes",
  "Mycoplasmatota",       "microbes",
  "Myxococcota",          "microbes",
  "Nitrospinota",         "microbes",
  "Nitrospirota",         "microbes",
  "Planctomycetota",      "microbes",
  "Pseudomonadota",       "microbes",
  "Rhodothermota",        "microbes",
  "Spirochaetota",        "microbes",
  "Synergistota",         "microbes",
  "Thermodesulfobacteriota", "microbes",
  "Thermomicrobiota",     "microbes",
  "Thermotogota",         "microbes",
  "Verrucomicrobiota",    "microbes",
  "Calditrichota",        "microbes",
  "Chrysiogenota",        "microbes",
  "Vulcanimicrobiota",    "microbes",
  "Elusimicrobiota",      "microbes",
  "Atribacterota",        "microbes",
  
  # fungi (only Oomycota - though technically fungus-like protists)
  "Oomycota",             "fungi",
  
  # protozoa
  "Ciliophora",           "protozoa",
  "Euglenozoa",           "protozoa",
  "Heterolobosea",        "protozoa",
  "Evosea",               "protozoa",
  
  # other (remaining protists/algae)
  "Bacillariophyta",      "other",
  NA,                     NA
)

# ------------------------------------------------------- ages -----------------------------------------------------
mw_ages <- read.csv("data/age_table.csv", header = TRUE, sep = ",") %>%
  rename(sample, library = sample)

#--------------------------------------------------- native vs non-native -----------------------------------------------
# Note this Only works for Plants. Other phyla will get a value, but this is irrelevant
mw_target_area <- c(
"GER" # Germany
)
#  look up which genera are native from WCVP
# exclude anything non-native form df, can be used to give indication of native
mw_wcvp_native <- wcvp_dist %>%
  filter(area_code_l3 %in% mw_target_area,
         introduced == 0) %>%
  right_join(
    wcvp_names %>% select(plant_name_id, genus),
    by = "plant_name_id"
  ) %>%
  select(genus, everything()) %>%
  filter(!is.na(genus),
         !is.na(area_code_l3)) %>%
  distinct(genus, region, area, area_code_l3)

# this just gives a more detailed overview which genus is native to which country
mw_wcvp_native_ov <- mw_wcvp_native %>%
  mutate(native = TRUE) %>%
  pivot_wider(
    names_from = area_code_l3,
    values_from = native,
    values_fill = FALSE,
    names_prefix = "native_"
  ) %>%
  mutate(native_any = TRUE)

#---------------------------------- assign location ------------------------------------------------------------
mw_site_assign <- tribble(
  ~library, ~site,
  "Lib.J.1038", "M",
  "Lib.J.1037", "M",
  "Lib.J.1036", "M",
  "Lib.J.1039", "W2",
  "Lib.J.1040", "W2",
  "Lib.J.1041", "W2",
  "Lib.J.1042", "W2",
  "Lib.J.1043", "W2",
  "Lib.J.1044", "W1",  
  "Lib.J.1045", "W1",
  "Lib.J.1046", "W1",
  "Lib.J.1047", "W1"
)



#------------------------------ add all to df's -----------------------------------------

datasets <- list(mw_2504, mw_2506, mw_2509)
names(datasets) <- c("mw_2504", "mw_2506", "mw_2509")

datasets <- lapply(datasets, function(df) {
  df %>%
    left_join(mw_phylum_groups, by = "phylum_name") %>%
      mutate(
    native_plant = case_when(genus_name %in% mw_wcvp_native$genus ~ "1",
                       TRUE ~ "0")
  ) %>%
    filter(library %in% c("Lib.J.1036", 
                          "Lib.J.1037", 
                          "Lib.J.1038", 
                          "Lib.J.1039", 
                          "Lib.J.1040", 
                          "Lib.J.1041", 
                          "Lib.J.1042", 
                          "Lib.J.1043", 
                          "Lib.J.1044", 
                          "Lib.J.1045", 
                          "Lib.J.1046", 
                          "Lib.J.1047")) %>%
    mutate(location_layer = case_when( # here we add location for each library
    str_detect(library, "Lib.J.1036")  ~ "ML3"  ,
    str_detect(library, "Lib.J.1037")  ~ "ML2"  ,
    str_detect(library, "Lib.J.1038")  ~ "ML1"  ,
    str_detect(library, "Lib.J.1039")  ~ "W2L4" ,
    str_detect(library, "Lib.J.1040")  ~ "W2L4" ,
    str_detect(library, "Lib.J.1041")  ~ "W2L3" ,
    str_detect(library, "Lib.J.1042")  ~ "W2L3" ,
    str_detect(library, "Lib.J.1043")  ~ "W2L3" ,
    str_detect(library, "Lib.J.1044")  ~ "W1L4" ,
    str_detect(library, "Lib.J.1045")  ~ "W1L4" ,
    str_detect(library, "Lib.J.1046")  ~ "W1L3_2",
    str_detect(library, "Lib.J.1047")  ~ "W1L3_1"  
  )) %>%
    left_join(mw_site_assign, by = "library") %>%
    left_join(mw_ages, by = "library") %>%
    select(library, location_layer, age, se_age, everything())
    
})

# unpack for normal usage, rename with _raw
mw_2504_raw <- datasets[["mw_2504"]]
mw_2506_raw <- datasets[["mw_2506"]]
mw_2509_raw <- datasets[["mw_2509"]]

```
# simplify df's
## define columns to keep
```{r, echo = FALSE}
# Define relevant columns for a genus df
mw_columns_df_fm_gs <- c(
  "library",
  "location_layer",
  "site",
  "age",
  "se_age",
  "kingdom_group",
  "phylum_name",
  "order_name",
  "family_name",
  "genus_name",
  "taxa_name", # taxa in this case refers to the reference genome which the read is mapped to, res is genus still
  "genus_count_assembly",
  "genus_deam_count",
  "sum_genus_count",
  "median_read_length_genus",
  "ancientness_genus",
  "deam5p_frac_genus",
  "deam5p_genus",
  "ref_C_sum_5p_genus",
  "deam3p_frac_genus",
  "deam3p_genus",
  "ref_C_sum_3p_genus",
  "family_count_assembly",
  "family_deam_count",
  "sum_family_count",
  "median_read_length_family",
  "ancientness_family",
  "deam5p_frac_family",
  "deam5p_family",
  "ref_C_sum_5p_family",
  "deam3p_frac_family",
  "deam3p_family",
  "ref_C_sum_3p_family",
  "native_plant"
  )

mw_columns_df_extra <- c(
  "library",
  "core",
  "depth",
  "age_ka",
  "db",
  "phylum_name",
  "order_name",
  "family_name",
  "genus_name",
  "taxa_name", # taxa in this case refers to the reference genome which the read is mapped to, res is genus still
  "genus_count_assembly",
  "sum_genus_count",
  "genus_deam_count",
  "median_read_length_genus",
  "ancientness_genus",
  "deam5p_frac_genus",
  "deam5p_genus",
  "ref_C_sum_5p_genus",
  "deam3p_frac_genus",
  "deam3p_genus",
  "ref_C_sum_3p_genus",
  "family_count_assembly",
  "family_deam_count",
  "sum_family_count",
  "median_read_length_family",
  "ancientness_family",
  "deam5p_frac_family",
  "deam5p_family",
  "ref_C_sum_5p_family",
  "deam3p_frac_family",
  "deam3p_family",
  "ref_C_sum_3p_family",
  "sum_phylum_count",
  "sum_order_count",
  "depth_min",
  "depth_max"
  )





#------------------ appply to df's ------------------------------------------------------------------
datasets2 <- list(mw_2504_raw, mw_2506_raw, mw_2509_raw)
names(datasets2) <- c("mw_2504_raw", "mw_2506_raw", "mw_2509_raw")

datasets2 <- lapply(datasets2, function(df2) {
  df2 %>%
  select(all_of(mw_columns_df_fm_gs)) %>% # use select to only select columns relevant for df_gs 
  group_by(library, genus_name) %>%
  mutate(
    deam5p_frac_genus_family = (deam5p_genus + deam5p_family) / (ref_C_sum_5p_genus + ref_C_sum_5p_family),
    deam3p_frac_genus_family = (deam3p_genus + deam3p_family) / (ref_C_sum_3p_genus + ref_C_sum_3p_family)) %>%
    filter(!genus_count_assembly < 50,
           !deam5p_frac_genus < 0.001)
})

# mw = Moehne Wester, then date, df = datafram, fm = family, gs = genus, cl = clean (filter for genus count > 100 and sum genus count > 100). all is just with all data

mw_2504_df_fm_gs_cl <- datasets2[["mw_2504_raw"]]
mw_2506_df_fm_gs_cl <- datasets2[["mw_2506_raw"]]
mw_2509_df_fm_gs_cl <- datasets2[["mw_2509_raw"]]
```


# median readlength vs deamination rate
try without collapsing to genus level first (include all assemblies)
```{r, echo = FALSE}
mw_colours <- c(
  "Lib.J.1038" = "#F07056",
  "Lib.J.1037" = "#E03612",
  "Lib.J.1036" = "#9E1C02",
  "Lib.J.1039" = "#74B4E8",
  "Lib.J.1040" = "#306FE3",
  "Lib.J.1041" = "#252FCF",
  "Lib.J.1042" = "#2E3494",
  "Lib.J.1043" = "#000373",
  "Lib.J.1044" = "#F781DF",  
  "Lib.J.1045" = "#D433B4",
  "Lib.J.1046" = "#8F0073",
  "Lib.J.1047" = "#5C0049"
)
mw_2509_raw_gs <- mw_2509_raw %>%
  filter(sum_genus_count > 1)
# define df per library for plotting (i know i should loop this but wtf)

mw_2509_df_M_cl <- mw_2509_df_fm_gs_cl %>%
  filter(site == "M")
mw_2509_df_M_di <- mw_2509_raw_gs %>%
  filter(site == "M")
mw_2509_df_W1_cl <- mw_2509_df_fm_gs_cl %>%
  filter(site == "W1")
mw_2509_df_W1_di <- mw_2509_raw_gs %>%
  filter(site == "W1")
mw_2509_df_W2_cl <- mw_2509_df_fm_gs_cl %>%
  filter(site == "W2")
mw_2509_df_W2_di <- mw_2509_raw_gs %>%
  filter(site == "W2")
#------------- Want to add number of observations per library to the plot--------------------
# clean one

# now n labels per fkn thing
mw_n_labels_M_cl <- mw_2509_df_M_cl %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
mw_n_labels_M_di <- mw_2509_df_M_di %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
mw_n_labels_W1_cl <- mw_2509_df_W1_cl %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
mw_n_labels_W1_di <- mw_2509_df_W1_di %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
mw_n_labels_W2_cl <- mw_2509_df_W2_cl %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
mw_n_labels_W2_di <- mw_2509_df_W2_di %>%
  group_by(library) %>%
  summarise(count = n()) %>%
  mutate(label = paste0("N = ", count))
# ----------------------- Plooooooooooooooooot - ------------------------------

mw_pres_plots_W2_cl <- ggplot(mw_2509_df_W2_cl, aes(x = deam5p_frac_genus, y = median_read_length_genus, fill = library)) +
  geom_point(alpha = 1, size = 3, shape = 21, colour = "black") +
  geom_smooth(colour = "red", alpha = 0.4, se = TRUE, level = 0.95, method = "lm", span = 0.9) +
  geom_text(data = mw_n_labels_W2_cl, aes(label = label), 
            x = 0.35, y = 75,  # adjust position as needed
            inherit.aes = FALSE)+
  scale_y_continuous(limits = c(30,80)) +
  scale_x_continuous(limits = c(0, 0.4)) +
  scale_colour_manual(values = mw_colours) +
  scale_fill_manual(values = mw_colours) +
  labs(
    title = "Wester 2: deamrate vs median readlength",
    x = "deam5p_frac_genus",
    y = "median_read_length_genus"
    ) +
  facet_wrap(~library, ncol = 3) +
  theme_classic() +
  theme(legend.position = "none")



mw_pres_plots_M_cl
mw_pres_plots_W1_cl
mw_pres_plots_W2_cl
mw_pres_plots_M_di
mw_pres_plots_W1_di
mw_pres_plots_W2_di


ggsave("output/mw_deamr_readlength_W2_clean.pdf", plot = mw_pres_plots_W2_cl, width = 12, height = 5, dpi = 300)


```


# --------------------- Plant and Fish lookup -------------------
```{r, echo = FALSE}

# ---------- FISH -----------------
mw_df_fish <- mw_2506_raw %>%
  filter(family_name %in% c("Cyprinidae", "Gobiidae", "Cichlidae", "Characidae", "Loricariidae", 
                            "Balitoridae", "Serranidae", "Labridae", "Scorpaenidae", "Castoridae")) %>%
  slice_max(genus_count_assembly, n = 1, with_ties = FALSE) %>%
  select(library, location_layer, age, phylum_name, order_name, family_name, genus_name, genus_count_assembly, sum_genus_count, deam5p_frac_genus, median_read_length_genus)

# ---------------- PLANTS -------------------
mw_df_plant <- mw_2509_df_fm_gs_cl %>%
  filter(kingdom_group == "plants") %>%
  group_by(library) %>%
  filter(!sum_genus_count < 100) %>%
  ungroup()
  
n_plant_genera <- mw_df_plant %>%
  summarise(
    n_pl_genera = n_distinct(genus_name)
  )

n_plant_genera

# ------------------ ANIMALS -------------------
mw_df_animal <- mw_2509_df_fm_gs_cl %>%
  filter(kingdom_group == "animals") %>%
  group_by(library) %>%
  filter(!sum_genus_count < 100) %>%
  ungroup()

n_animal_genera <- mw_df_animal %>%
  summarise(
    n_animal_genera = n_distinct(genus_name)
  )
n_animal_genera

mw_bovidae <- mw_df_animal %>%
  filter(family_name == "Bovidae") 
```
